Magalhaes and Schwilk Supplementary Materials

### 2019-05-13 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Add di and maxmc trait models

%=================================================================
% Full title of the paper (Capitalized)
\Title{Moisture effects on leaf litter are species-specific and result in non-additive effects flammability in mixed conifer forest}

@@

% Authors, for the paper (add full first names)
\Author{Rita Quinones de Magalhaes $^{1,\dagger,\ddagger}$\orcidA{} and Dylan W. Schwilk $^{2,\ddagger}$}

% Authors, for metadata in PDF
\AuthorNames{Rita Magalhaes, Dylan W. Schwilk}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
$^{2}$ \quad Texas Tech University; dylan.schwilk@ttu.edu}

@@

% Current address and/or shared authorship
\firstnote{Current address: Affiliation 3} 
\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

@@

We tested if litter bulk density influenced moisture dynamics (maximum water retainability and desiccation rate) by fitting linear models with species means as observations.

@@

@@

\begin{figure}[H]
  \centering
  \label{fig-drydown}
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption{Dry down curves for eight litter types. Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis. Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing. Colors indicate genera and there were significant differences in both }
\end{figure}

@@

Across these eight species litter bulk density had a negative effect on desiccation rate (Fig. \ref{fig-bd-di}, linear model p = 0.011).

\begin{figure}[H]
  \centering
  \label{fig-bd-di}
\includegraphics[width=8cm]{figs_tables/fig3_di_by_bd.pdf}
\caption{ }
\end{figure}

\subsection{Time since wetting and its effect on flammability}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Subsection}
\unskip
\subsubsection{Subsubsection}

Bulleted lists look like this:
\begin{itemize}[leftmargin=*,labelsep=5.8mm]
\item	First bullet
\item	Second bullet
\item	Third bullet
\end{itemize}

Numbered lists can be added as follows:
\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
\item	First item 
\item	Second item
\item	Third item
\end{enumerate}

The text continues here.

\subsection{Figures, Tables and Schemes}

All figures and tables should be cited in the main text as Figure 1, Table 1, etc.

Text

Text

@@

The flammability results for the mixtures followed closely the results obtained for the individual species. Mixtures containing Q. kelloggii and P. jeffreyi had faster spread rates and lower times to ignition, resulting in higher overall flammability. These results are influenced by the lower bulk densities of litter mixtures containing those two species. There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter \citep{Magalhaes+Schwilk-2012, VanAltena+Logtestjin+etal-2012, }. Spread rate and time to ignition had significant non-additivity, and were strongly influenced by the species present in the mixture. Presence of C. decurrens in a mixture marked a big difference between the null model and the prediction. This was due to several failed ignitions for this species when burned individually, which conditioned the predicted values, while all the mixtures ignited. There were marked differences in the non-additivity results from this study and the results from a previous study investigating dry litter samples. The presence of non-additive effects in moisture content leads to non-additive effects in the flammability of moist litter beds. When accounting for this effect, the non-additivity on the flammability of moist litter beds is subtler, when compared to species effects. This result, together with previously published work, suggests that litter flammability is prone to non-additive effects, reinforcing the idea that we cannot approach flammability by lumping species into fuel categories, as is common in fire behavior models. Community flammability is dependent on and particular to its constituent species and cannot be addressed by simply averaging the flammability of all species it comprises. This dynamic view is likely to be increasingly relevant considering ongoing climate change, which predicts changes in community assembly.

I have found non-additive effects (both positive and negative) in moisture content and in the flammability of wet litter mixtures of mixed-conifer forest. These results complement the efforts of Blauw et al. (2015) for moist litter and add to the literature investigating non-additivity in litter flammability (de Magalhães and Schwilk 2012, van Altena et al. 2012). However, these studies are limited in the scope of the species studied and suggest that non-additivity needs to be investigated at a larger scale, within different systems and comparatively across a larger number of species. Moisture content is considered the single most important driver of litter flammability and the results from my study indicate how leaf and litter traits affect moisture dynamics, and how moisture effects on litter flammability are species-specific. A trait-based perspective provides a better mechanistic understanding of moisture dynamics of litter fuel flammability and fire behavior and is especially important for predicting and understanding fire in plant communities that are changing.

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest}

@@

% Authors, for the paper (add full first names)
\Author{Rita Quinones de Magalhaes $^{\dagger}$ and Dylan W. Schwilk}

% Authors, for metadata in PDF
\AuthorNames{Rita Magalh\~{a}es, Dylan W. Schwilk}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{Texas Tech University}

@@

% Current address and/or shared authorship
\firstnote{Current address: Rochester Institute of Technology, Rochester, NY 14623} 
%\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

@@

We tested if one leaf trait (specific leaf area) and one litter trait (litter
bulk density) influenced moisture dynamics (maximum water retainability and
desiccation rate) by fitting linear models with species means as observations.
For each response variable we fit a single linear model with two predictors and
an interaction term.

@@

\subsection{Species specific moisture dynamics and leaf traits}

@@

\begin{figure}[H]
  \centering
  \label{fig1-drydown}
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption[Dry down curves for eight litter types.]{Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis. Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing. Colors indicate genera and there were significant differences in both }
\end{figure}

@@

Across these eight species specific leaf area had a positive effect on maximum water retention (Fig. \ref{maxmc}, linear model p = 0.003 and litter bulk density had a negative effect on desiccation rate (Fig. \ref{fig-bd-di}, linear model p = 0.029). Anova tables are shown in in Fig. \ref{tab-mc-anova} and Fig. \ref{tab-di-anova}.

\begin{figure}[H]
  \centering
  \label{fig-maxmc-di}
\includegraphics[width=16cm]{figs_tables/fig3-SLA-maxMC.pdf}
\caption[Maximum water retention by specific leaf area]{Maximum water retention by specific leaf area (SLA) across eight species.  TODO }
\end{figure}

\begin{table}[H]
  \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density.}
  \label{tab2-mc-anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/mc-anova.ltx}
\end{table}

\begin{figure}[H]
  \centering
  \label{fig-bd-di}
\includegraphics[width=16cm]{figs_tables/fig4_di_by_bd.pdf}
\caption[Litter density effect on dessication rate.]{  TODO }
\end{figure}

\begin{table}[H]
  \caption{Linear model results for estimated dessication rate ($hr^{-1}$) as a function of specific leaf area and litter bulk density.}
  \label{tab2-di-anova}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/di-anova.ltx}
\end{table}

\subsection{Time since wetting effect on flammability}

@@

@@

The flammability results for the mixtures followed closely the results obtained for the individual species. Mixtures containing Q. kelloggii and P. jeffreyi had faster spread rates and lower times to ignition, resulting in higher overall flammability. These results are influenced by the lower bulk densities of litter mixtures containing those two species. There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter \citep{Magalhaes+Schwilk-2012, VanAltena+Logtestjin+etal-2012}. Spread rate and time to ignition had significant non-additivity, and were strongly influenced by the species present in the mixture. Presence of C. decurrens in a mixture marked a big difference between the null model and the prediction. This was due to several failed ignitions for this species when burned individually, which conditioned the predicted values, while all the mixtures ignited. There were marked differences in the non-additivity results from this study and the results from a previous study investigating dry litter samples. The presence of non-additive effects in moisture content leads to non-additive effects in the flammability of moist litter beds. When accounting for this effect, the non-additivity on the flammability of moist litter beds is subtler, when compared to species effects. This result, together with previously published work, suggests that litter flammability is prone to non-additive effects, reinforcing the idea that we cannot approach flammability by lumping species into fuel categories, as is common in fire behavior models. Community flammability is dependent on and particular to its constituent species and cannot be addressed by simply averaging the flammability of all species it comprises. This dynamic view is likely to be increasingly relevant considering ongoing climate change, which predicts changes in community assembly.

% I have found non-additive effects (both positive and negative) in moisture content and in the flammability of wet litter mixtures of mixed-conifer forest. These results complement the efforts of Blauw et al. (2015) for moist litter and add to the literature investigating non-additivity in litter flammability (de Magalhães and Schwilk 2012, van Altena et al. 2012). However, these studies are limited in the scope of the species studied and suggest that non-additivity needs to be investigated at a larger scale, within different systems and comparatively across a larger number of species. Moisture content is considered the single most important driver of litter flammability and the results from my study indicate how leaf and litter traits affect moisture dynamics, and how moisture effects on litter flammability are species-specific. A trait-based perspective provides a better mechanistic understanding of moisture dynamics of litter fuel flammability and fire behavior and is especially important for predicting and understanding fire in plant communities that are changing.

### 2019-05-13 ms/SI.tex
Add di and maxmc trait models

  \includegraphics[width=17cm]{figs_tables/figS1_max_water_emmeans.pdf}
\caption{Estimated marginal means by species for maximum water retainability based on mixed model results.}

@@

\begin{figure}[h]
  \centering
  \label{fig-S2}
\includegraphics[width=17cm]{figs_tables/figS2_dessication_emmeans.pdf}
\caption{Estimated marginal means by species for dessication rate based on mixed model results.}
\end{figure}

>>>>

  \includegraphics[width=17cm]{figs_tables/figS1_max_water_emmeans.pdf}
\caption{Estimated marginal means by species for maximum water retainability based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}

@@

\begin{figure}[h]
  \centering
  \label{fig-S2}
\includegraphics[width=17cm]{figs_tables/figS2_dessication_emmeans.pdf}
\caption{Estimated marginal means by species for dessication rate based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
\end{figure}
